---
title: "HW1_S23"
author: "Nicole Higgins"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
options(scipen = 1, digits = 16)
```


## Homework 1, due Wednesday 1/25 at 5PM 
Homework guidelines: 

- Turn in this assignment as a PDF on Moodle, please! To create a PDF, I suggest knitting to HTML, then opening the HTML file in a browser and saving to PDF from there. 

- You're invited to collaborate and discuss with each other, and then each person should turn in their own assignment, which should be their own work.  "Discussing" is very different from copying and I trust students to stay on the right side of this line. In general, anything you say out loud to another person is fine, and looking at a screen together (in person or on Zoom) is fine. Sharing files or screenshots is a bad idea. Name the people you work with below (question 0). 

- If you start early, you're giving yourself the chance to ask questions and turn in a polished product. If you start late, you won't be able to get help as easily. 

### Q0: Names
Who did you work with on this assignment? You'll get a bonus 5 points if you name someone and they name you. *No scores above 100% are allowed, but these bonus points can repair mistakes on this assignment*.

**I worked with Jonny :)**

### Q1: Eigenvalues of a symmetric matrix

- Suppose that $A$ is an $n\times n$ matrix and that $x$ is a vector in $\mathbb{R}^n$. Is $x^TAx$ a number, vector, or matrix? 

**it is a number, since the dimensions [1 x n][n x n][n x 1] = [1 x 1]**

- Suppose that $Ax = \lambda x$ and $A = A^T$ and $x$ is not all zeros; that is, $\lambda$ is an eigenvector of a symmetric matrix $A$ with eigenvalue $\lambda$. Show that $x^TAx = \lambda x^Tx$. Explain why this means that $\lambda$ must be a real number. 

**Given $Ax = \lambda x$, then $x^TAx = x^T\lambda x$. Since $\lambda$ is a scalar, $x^T\lambda x = \lambda x^Tx$. Solving for $\lambda$, we get $\lambda = x^TAx/x^Tx$. Given that $x$ is not all zeros, we know $x^T$ is also not all zeros. Thus, $x^Tx$ is nonzero so $\lambda$ is real.**

- Form the matrices $P$ and $S$ in R and use the syntax *eigen(A)$values* to find their eigenvalues. (According to the result we proved, $S$ should have real eigenvalues and $P$ might or might not have real eigenvalues). 
$$P = \begin{pmatrix} -4&5&6\\1&0&9 \\9&1&2\end{pmatrix}
\qquad
S = \begin{pmatrix} -1&2&1\\2&7&-3\\1&-3&4\end{pmatrix}$$

```{r Question 1.3}
pData = c(-4,1,9,5,0,1,6,9,2)
sData = c(-1,2,1,2,7,-3,1,-3,4)

P = matrix(pData, 3, 3)
S = matrix(sData, 3, 3)

print(P)
print(S)

eigen(P)$values
eigen(S)$values
```

### Q2: Nearly R-O
Write some code to create and print out this matrix (it's possible to do it in two lines, but any solution is fine as long as you don't separately type all 16 entries). Explain why it is nearly a R-O matrix, where "R-O" stands for something on our first-day slides. 
$$ 
A = \begin{pmatrix}
1&3&5&10\\
10&30&50&100\\
-2&-6&-10&-20\\
100&300&4&4
\end{pmatrix}
$$

```{r}
A <- c(1, 10, -2, 100) %*% t(c(1, 3, 5, 10))
A[4,3:4] <- 4

print(A)
```


### Q3: Basic graph example
Write some code to plot the quadratic function $y=x^2 - 5x + 2$. Make sure that the minimum is visible inside your figure.  

```{r}
x = -100:100
y = x^2 - 5*x + 2

plot(x,y)
```

### Q4: When power functions look linear
Consider the three graphs that the following code produces. *Note: I don't need to see the graphs in the version you turn in.  Please put "echo=FALSE" on line 60 once you finish looking at the graphs.*
```{r, echo=FALSE, eval=FALSE}
x = 1:1000
y = x^(-2.0)
plot(x,y)
plot(log10(x), log10(y))
plot(x,y,log="xy")
```

Comment carefully on the differences you see between the three versions. Then explain why a function of the form $y = x^p$ looks linear on a log-log plot. *Hint: let $w = \log_{10}(y)$ and let $t = \log_{10}(x)$, then show that there is a linear relationship between $t$ and $w$.*

**The first plot is on traditionally scaled axes - that is, x and y increment by one each tick - and the resulting graph is nearly flat and difficult to interpret. The second plot scales the axes as logarithmic**

### Q5: Create a letter
As on the Day 1 activity, create a color image showing the first letter of your name by building appropriate matrices in R. Give it a white background (initialize the matrices with 1's instead of 0's). Add it to the Google Doc, and also include it here.

### Q6: High contrast image
The image "Department.png" is a cropped version of a photo by Stefani Reynolds in a January 2023 article about the upcoming US debt ceiling crisis in the New York Times. It shows part of the Treasury Department building in Washington. It is a color image, but it looks pretty gray because for most pixels, the values in the R, G, and B channels are all close together. Save the image in the same folder as your main RMD file and then uncomment these lines to see it: 

```{r}
#library("magick")
#a = image_read("Department.png")
#A = as.integer(a[[1]])/256
#redA = A[,,1]
#greenA = A[,,2]
#blueA = A[,,3]
#print(a) # this displays the image in RStudio and knitted documents
```

We will make a high-contrast version of this image using a new code trick: updating a matrix or vector using conditional tests. 

Here is a first example where we create a vector and then ask which entries are less than 5:
```{r}
v <- 1:8
print(v)
print(v < 5)
```
The useful trick is that you can update the entries for which the condition is true: 
```{r}
v[v<5] <- -1
print(v)
```

The same trick works for matrices and high-dimensional arrays: 
```{r}
a <- matrix(rnorm(12),3,4) # create a 3x4 matrix with normally distributed entries
a[a<0] <- 0 # set all negative elements to be zero
print(a) # we should see about half zeros, half positive numbers
```

The three matrices redA, greenA, and blueA contain numbers between 0 and 1. Pick a cutoff between 0 and 1 (0.5 is a good choice but you can experiment) and then set all entries to 1 if they're greater than the cutoff, or 0 otherwise. Then display the altered image, the "high contrast" version. It should be mostly black-and-white but there will be some colored pixels - explain why. 




