---
title: "Givens + Householder Activity"
author: "Nicole Higgins"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## $A = QR$ by Givens rotations

```{r}
# Plan: staring with square A, repeatedly apply Givens rotations:
#             GGGGGGGGA = R 
# (put in zeros until we get a upper-triangular product)
# then, A = QR with Q = (GGGGGGG)^T
A <- matrix(runif(25),5,5)
A.original <- A + 0

n <- dim(A)[1]

Q <- diag(1,n,n)
for (co in 1:(n-1)){
  for (ro in n:(co+1)){
    # put in a zero at position (ro,co) 
    # and dump the extra mass into the diagonal entry (co,co) 
    
    a <- A[co,co]
    b <- A[ro,co]
    r <- sqrt(a^2+b^2)
    c <- a/r
    s <- -b/r
    G <- diag(1,n,n)
    G[co,co] <- c
    G[ro,ro] <- c
    G[ro,co] <- -s
    G[co,ro] <- s
    A <- G %*% A
    Q <- Q %*% t(G)
        
    
  }
}
R <- A

print(max(abs(A.original - Q %*% R  )))

```



## $A = QR$ by Householder Reflections
This is how R's built-in QR factorization method works! 

```{r}
# Plan: multiply A by Householder reflections H until HHHHHHHHHH A = R is upper triangular
# then A = (H^TH^TH^TH^TH^TH^TH^TH^T) R
# or A = Q R

# assume A is square

A <- matrix(runif(25),5,5)
A.original <- A + 0 
n <- dim(A)[1]
Q <- diag(1,n,n)

for (co in 1:(n-1)){
  # put 0s below the diagonal in column "co" of A
  x <- A[,co] # column co of A
  w <- x + 0
  w[(co+1):n] <- 0
  w[co] <- sqrt(sum(x[co:n]^2))
  
  #reflector
  v <- (x - w)
  nv <- sqrt(sum(v^2))
  v <- v/nv
  H <- diag(1,n,n) - 2 * v %*% t(v)
  
  #reflecting
  A <- H %*% A
  Q <- Q %*% t(H)
}

R <- A

print(R)
print(max(abs(A.original - Q %*% R)))
print(max(abs(diag(1,n,n) - t(Q) %*% Q)))
```


## $A = U\Sigma V^T$ by Givens rotations
In this context, the Givens rotations are called Jacobi rotations :) 


#### A long warmup question
Previously we used Givens rotations to put zeros into vectors or matrices. Now we'll use them for another purpose: making two vectors perpendicular. 

Suppose that $A$ is a $2\times n$ matrix with rows $A_1^T$ and $A_2^T$.  If we make $GA=N$ where $G$ is the Givens rotation $G=\begin{pmatrix}c&-s\\s&c\end{pmatrix}$, then the new rows will be $N_1^T = cA_1^T-sA_2^T$ and $N_2^T = sA_1^T + cA_2^T$. Can you find a way to pick $s$ and $c$ so that $0 = N_1^T N_2 = sc (A_1^TA_1 - A_2^TA_2) +(c^2-s^2)A_1^TA_2$?

Solution: use $\cos(2t) = \cos^2(t) - \sin^2(t)$ and $\sin(2t) =2 \sin(t)\cos(t)$ to get
$$\cos(2t) A_1^TA_2 = \frac12 \sin(2t)(A_2^TA_2-A_1^TA_1) $$ 
If $A_1^TA_2=0$ we can take $t=0$: the rows were already perpendicular and we don't to rotate.
If $A_2^TA_2-A_1^TA_1=0$ then we can take $t = \pi/4$. 
Otherwise, we can divide to get 
$$ 
\tan(2t) = \frac{2A_1^TA_2}{A_2^TA_2-A_1^TA_1}
$$
But now we can use the formula $\tan(2t) = 2\tan t / (1-\tan^2 t)$ to get 
$$
\tan t = (1-\tan^2 t)\frac{A_1^TA_2}{A_2^TA_2-A_1^TA_1}
$$
With $\beta = \frac{A_2^TA_2-A_1^TA_1}{2A_1^TA_2}$, this becomes the quadratic: 
$$
\tan^2 t + 2\beta\tan t - 1 = 0
$$
The solutions of the quadratic are
$$
\tan t =  {-\beta} \pm \sqrt{\beta^2+1}
$$
To avoid subtracting nearly equal quantities, we pick one:
$$
\tan t =  -\sgn(\beta)\left(|\beta|- \sqrt{\beta^2+1}\right) =  -\sgn(\beta)\frac{-1}{|\beta|+ \sqrt{\beta^2+1}} = \frac{\sgn(\beta)}{|\beta|+ \sqrt{\beta^2+1}}
$$
Finally, get $c = 1/\sqrt{\tan(t)^2 + 1}$ and $s = \tan(t) c$.

#### SVD by left Jacobi
Multiply $GGGGGGA$ until the rows of $A$ are orthogonal, but not unit length. Then let $\sigma_i $ be the $2$-norm of row $i$. Then $GGGGGGA = \Sigma V^T$ and $A = (GGGGGG)^T \Sigma V^T$. 




