---
title: "Polynomial and Rational Interpolation"
author: "Nicole Higgins"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Singular values of Vandermonde
Create some vector of 20 $x$-values and then make a $20\times20$ Vandermonde matrix. What are the singular values? What is the condition number $\sigma_1 / \sigma_{20}$? Recall that you will lose $\log_{10}(\sigma_1/\sigma_n)$ digits of accuracy when solving $Ax=b$. Try other sizes for your Vandermonde matrix and explore the results. 

```{r}
xs <- c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20)

vand <- outer(xs,0:19,'^')

svd(vand)$d[1]/svd(vand)$d[20]

log10(svd(vand)$d[1]/svd(vand)$d[20])
```



### Graph a polynomial 
Before you type anything, make a sketch of the graph by hand! 
$$y = \frac{(x-1)(x-2)(x-3)(x-5)}{(4-1)(4-2)(4-3)(4-5)} $$


### A polynomial approximation of $e^x$ 
Plot polynomial of degree $\le 6$ that passes through these points: 
```{r}
x <- cos(((1:7) - 1)*pi/6)
y <- exp(x)
plot(x,y)
```




### Draw the moon
Here are some dots for the outline of a crescent moon. Use rational interpolation to draw in the full outline of the moon! 
```{r}
moon.data <- rbind(c(798, 150), c(417, 382), c(133, 760), c(30, 1198), c(89, 1639), c(354, 
  2065), c(751, 2341), c(1409, 2414), c(1839, 2239), c(1203, 2029), c(903, 
  1707), c(730, 1255), c(758, 755), c(962, 353), c(1262, 76))
x <- moon.data[,1]
y <- moon.data[,2]
plot(x,y,asp=1)
```

