---
title: "SVD + Interpolative + Radio"
author: "Nicole Higgins"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Activity: Using SVD, Interpolative decomposition for radio signals


#### E1: Make random sources
Create 120 random points inside the box $[6,7]\times[4,5]$.  Make a plot showing them. 

```{r}
e1x <- runif(120, 6, 7)
e1y <- runif(120, 4, 5)

plot(e1x, e1y)
```



#### E2: Make random targets
Create 100 random points inside the box $[0,1]\times[0,1]$.  Make a plot showing them. 

```{r}
e2x <- runif(100, 0, 1)
e2y <- runif(100, 0, 1)

plot(e2x, e2y)
```


#### E3: Evaluate the signal at one target. 
Print out the coordinates of the first target. Call these $x_1$ and $y_1$. 
Let the coordinates of the sources be $(X_j,Y_j)$ where $1\le j \le 120$. 
If each source is emitting a signal of strength $f_j$, then the first target hears a combined signal of strength 
$$s_1 = \sum_{j=1}^{120} \frac{f_j }{\sqrt{(x-X_j)^2+(y-Y_j)^2}}$$. 
Note that the total strength will be weaker if the sources are far from the target. 

Suppose that the strength of source $j$ is the number $100/\sqrt{j+2}$.  Then find the strength of the signal perceived by the first target - that is, evaluate $s_1$.  

```{r}
s1x <- e2x[1]
s1y <- e2y[1]


s1 <- 0
for(j in 1:120){
  strength <- 100/sqrt(j+2)
  jsum <- strength/sqrt((s1x-e1x[j])^2+(s1y-e1y[j])^2)
  s1 <- s1 + jsum
}
s1
```



#### E4: Evaluate the signal at all targets. 

For all $i = 1,\cdots,100$ evaluate 
$$s_i = \sum_{j=1}^{120} \frac{f_j }{\sqrt{(x_i-X_j)^2+(y_i-Y_j)^2}}.$$
To do this, you should create a matrix $A$ with $A_{ij} = 1/\sqrt{(x-X_j)^2+(y-Y_j)^2}$ and then multiply by $\vec f$.  


```{r}
# strength vector
f <- matrix(0, 120, 1)

for(j in 1:120){
  f[j] <- 100/sqrt(j+2)
}

# matrix A
A <- matrix(0, 100, 120)

for(i in 1:100){
  for(j in 1:120){
    A[i,j] <- 1/sqrt((e2x[i]-e1x[j])^2+(e2y[i]-e1y[j])^2)
  }
}

# result
A %*% f
```



#### E5 SVD approximation
Find the best rank-one approximation of your matrix $A$ using the SVD and use it to approximate the answers to parts E3 and E4.
What is the numerical rank of your matrix? 


```{r}
out <- svd(A)
U <- out$u
sigma <- out$d
V <- out$v

aprx <- sigma[1] * (U[,1] %*% t(V[,1]))
sigma[1]
```



#### E6 Interpolative decomposition
Find the rank-$k$ interpolative decomposition of your matrix $A$ for $k=15$ or $k=7$ or $k=4$ and use it to approximate the answers to parts E3 and E4. How do you interpret this factorization? Which 15 columns get picked? 


```{r}
library('rsvd')
id <- rid(A, 9)
C <- id$C
Z <- id$Z
print(norm(A, "F"))
print(norm(A - C %*% Z, "F"))
print(id$idx)
```





