# Least-squares demo problems
 
### 1: Quadratic fit by guessing
Given the data set $(1,1)$, $(2,10)$, $(3,9)$, $(4,16)$, we can try to find the best-fit quadratic. Let $f(t) = at^2+bt+c$. This leads to $A\vec x=\vec b$ where $A$ is $4\times 3$. Let $g(x) = \|A\vec x-\vec b\|^2$. What is $g(x)$ if $x=(2,2,2)$, that is, if $a=b=c=2$? 
Then write some code that starts with a guess for $x$ and computes $g(x)$ and plots the quadratic with the four data points. How low can you make $g(x)$ just by random guessing or educated guessing based on the graphs? 

#### Solution
```{r}
A <- cbind(c(1,1,1,1),c(1,2,3,4),c(1,4,9,16))
b <- c(1,10,9,16)
g <- function(x){
  v <- A %*% x - b
  return(sum(v^2)) # equivalent: return(t(v) %*% v)
}
print(g(c(2,2,2))) # residual squared for a=b=c=2
```

```{r}
xg <- c(1,3,0) # a guess for x 
plot(c(1,2,3,4),b,type="p",main="A quadratic fit?")
t <- seq(0.9,4.1,length=400)
y <- xg[1] + t * xg[2] + t^2 * xg[3]
lines(t,y)
print(g(xg)) # you can do better by modifying xg! 
```

### 2: Quadratic fit by $Rx = Q^Tb$
Solve the previous question via least squares using $A=QR$ and then $Rx = Q^Tb$.  How low does $g(x)$ become for this official answer? 

#### Solution 1

```{r}
# Solution 1, using the built-in one line version:
x.1 <- qr.solve(A,b,tol=1e-14)
print(x.1)
print(g(x.1)) # this should be the smallest possible value of g
plot(c(1,2,3,4),b,type="p",main="Best quadratic")
t <- seq(0.9,4.1,length=400)
y <- x.1[1] + t * x.1[2] + t^2 * x.1[3]
lines(t,y)
```

#### Solution 2

```{r}
# solution 2, showing more explicit steps
QRT <- qr(A)
Q <- qr.Q(QRT)
R <- QRT$qr
R[lower.tri(R)] = 0.0
# Solve system
x.2 <- backsolve(R,t(Q) %*% b)
print(x.2)
print(g(x.2)) # this should be the smallest possible value of g
plot(c(1,2,3,4),b,type="p",main="Best quadratic")
t <- seq(0.9,4.1,length=400)
y <- x.2[1] + t * x.2[2] + t^2 * x.2[3]
lines(t,y)


```


### 3: Linear fitting by $Rx=Q^Tb$
Find the numbers $c_1$ and $c_2$ so that the line $y = c_1 + c_2x$ is the best possible fit to the data points $(1,7.55)$, $(2,8.4)$, $(3,6.35)$, $(4,0.2)$ in the least squares sense. Then plot the line and the points. 

#### Solution
```{r}
# enter A and b
A <- cbind(c(1,1,1,1),c(1,2,3,4))
b <- c(7.55,8.4,6.35,0.2)
# factor A = QR
QRT <- qr(A)
Q <- qr.Q(QRT)
R <- QRT$qr
R[lower.tri(R)] = 0.0
# Solve system
x <- backsolve(R,t(Q) %*% b)
# plot
plot(0:5,x[1]+x[2]*(0:5),'l',xlab="x",ylab="y",main="Least squares fit line")
points(1:4,b,pch=20,col="blue")

```





