---
title: "Code & practice questions: A=QR"
author: "(name)"
date: "3/28/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Here's an implementation of the $A=QR$ factorization procedure following the Gram-Schmidt idea. 
```{r}
my.QR <- function(A){
  # This is the numerically stable (good) version 
  # Also known as "modified Gram-Schmidt" or "mgs"
  # The input A must be square and invertible
  d <- dim(A)
  n <- d[1] # the matrix A is n-by-n
  Q <- 0*A  # Start with a Q full of zeros
  R <- 0*A  # Start with an R full of zeros
  my.len <- function(v){return(sqrt(sum(v^2)))} 
  for (i in 1:(n-1)){ # loop over rows of R, except the last 
    R[i,i] <- my.len(A[,i])  # find diagonal entry in R
    Q[,i] <- A[,i] / R[i,i]  # find i-th column of Q
    for (j in (i+1):n){
      R[i,j] <- sum(Q[,i] * A[,j]) # fill in the row in R
      A[,j] <- A[,j] - R[i,j]*Q[,i] # modify col of A
      # Previous line is the "modified" part of the algorithm
      # We overwrite cols of A as we go. 
    }
  }
  R[n,n] <- my.len(A[,n]) # get the last diagonal entry of R 
  Q[,n] <- A[,n] / R[n,n] # get the last column of Q
  return(list(R=R,Q=Q))
}
```

Here is an example showing how to use our implementation: 
```{r}
A <- matrix(runif(25),nrow=5)
myQRthing <- my.QR(A)
R <- myQRthing$R
Q <- myQRthing$Q
print(max(abs(A - Q %*% R))) # should be 0 if A = QR
```

Here is an example showing how to use R's built-in implementation. If you want to see the code, it's [here](https://svn.r-project.org/R/trunk/src/appl/dqrdc2.f) (rabbit hole warning). 
```{r}
A <- matrix(runif(25),nrow=5)
QRthing <- qr(A)
R <- QRthing$qr
R[lower.tri(R)] = 0.0  
Q <- qr.Q(QRthing)
print(max(abs(A - Q %*% R))) # should be 0 if A = QR
```

# Practice problems

### 1: Solve $Ax=b$ by making it into $QRx=b$
Here is a matrix $A$ and a vector $b$. 
Use our code, or the built-in R version, to find $x$ so that $Ax=b$. Confirm your answer by showing that all entries in $r=Ab-x$ are small. 
```{r}
A <- c(44,-72,205,565,-10,24,-10,423,19,-82,159,198,2,82,-68,-191)
dim(A) <- c(4,4)
A <- t(A) 
b <- c(66,419,-178,-67)
# First, factor A = QR:

# Next, solve for x: 

# Finally, check that Ax = b:

```
*Note: there is a built-in function that does all of this at once. You can type "x = solve.qr(A,b)" to go directly from $A$ and $b$ to $x$. It is even a little faster than if you first form $Q$ and $R$.*

### 2: Try $A=QR$ by hand
Try this on paper to make sure you understand how the Gram-Schmidt procedure works. Feel free to use a calculator or the R command line to make the arithmetic easier. I did rig it so that the numbers should be kind of nice. AFTER you finish :) check it by machine. 
$$A = \begin{pmatrix}
-80&-148&150\\0&-15&-100\\60&136&-50
\end{pmatrix}$$



### 3. What if $A$ isn't invertible? 
Here is a $3\times3$ matrix $A$ whose last column is the sum of the first two: 
$$A = \begin{pmatrix}
1&-1&0\\
3&2&5\\
-1&4&3
\end{pmatrix}$$

IRL, hopefully you would know in advance that you are working with a noninvertible matrix $A$, and you would not try to solve $Ax=b$ in that situation (instead you might use least squares, which is coming up later). It is interesting, though, to think about what would happen if you try to write $A=QR$ when $A$ is noninvertible. 

* What happens if you use our implementation of $A=QR$?
* What happens if you use the built-in $A=QR$? 
* Mathematically, what would go wrong if you try to do $A=QR$ on paper here?  How does the built-in $A=QR$ address this problem? 

### 4. Timing
Let's see how large a matrix $A$ you could factor into $QR$ if you had an hour of computing time on your machine. 
One way to do this is to check the time before and after you run the code. For example, here is how long it takes to add up the first $10^6$ integers using a for-loop:
```{r}
tstart = Sys.time()
t = 0.0
for (j in 1:1000000){
  t <- t + j
}
tend <- Sys.time()
print(tend - tstart)
```
It is much faster to create a vector and sum the entries, avoiding the for-loop! Example: 
```{r}
tstart = Sys.time()
t <- sum(1:1000000)
tend <- Sys.time()
print(tend - tstart)
```
* Make a random $100\times100$ matrix $A$ and time the factorization $A = QR$ using our code and the built-in one. Which is faster?  
* Now try a $200\times200$ matrix and a $400\times400$ matrix. Use the built-in code. Complete the sentence: "When I double the dimension of $A$, the computation time for $A = QR$ gets multiplied by approximately a factor of ???".  
* An hour is 3600 seconds. Predict how large a matrix you can factor in one hour on your machine. 


